---
title: "Team 12 Case Study"
author: "Lisa Zuo, Gian Castillero, Jack Nowacek"
date: "2025-04-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(tidyverse)
library(knitr)
library(naniar)
library(gridExtra)
library(GGally)

forcedeck <- read_excel("data/forcedeck.xlsx")
forceframe <- read_excel("data/forceframe.xlsx")
nordbord <- read_excel("data/nordbord.xlsx")
perch <- read_excel("data/perch.xlsx")
```

## Introduction

Duke Football is interested in using biomechanics data to improve performance and prevent injury in athletes. Using four instruments, they have collected physiometric data from athletes on the football team over the course of roughly a year. Athletes used each of these intruments multiple times over the course of that period.

The four instruments used to collect biomechanical data are `forcedeck`, `forceframe`, `nordbord`, and `perch`. The `forcedeck` instrument measures the power, acceleration, and other mechanics generated from the athlete's lower body when the athlete jumps straight up while their hands are on their hips. The `forceframe` instrument asseses the isometric strength of an athlete's different joints, especially hips, by measuring the force generated from the athlete's legs when they push their legs in or out while lying on their back. The `nordbord` instrument calculates hamstring stregth by measuring the force generated from the lower legs when the athlete lowers their body from a kneeling position. The `perch` instrument measures the power and velocity generated by the athlete while performing a squat with weights. Data from the football players were collected from these instruments and provided to the statistical team as four separate datasets.

We focused our analysis on two research questions. First, we investigated whether performance metrics from three of the instruments could be used to find associations with performance on a fourth instrument. Second, we examined whether asymmetries in performance (differences between left and right limbs) recorded on one instrument were associated with performance on another instrument. Understanding these two types of associations can not only improve the physical preparation of athletes and their performance but also identify weaknesses that affect performance in other, seemingly unrelated body parts, and prevent injuries.

The research described above was motivated by understanding that each of these test individually measures a single body part but an athlete needs its full body in peak form in order to perform at the highest level. Therefore, athletes preparation for elite competition can increase in efficiency if there exists body parts which increased performance is associated with increased performance in others. In other words, we intend to find a way to increase overall body performance from a increases in performance in a single or couple body parts. The motivation for investigating asymmetries is very similar. We intend to understand the association between imbalances in one body part and performance in others. An association between these might indicate that in a single body part might be associated to decrease in overall performance and not only in a specific test. This is also important in athletes preparation for high level competition. 

Exploratory plots of the data show that the data appear to be right-skewed with significant outliers.

```{r include=FALSE}
# Select blue and green columns
forcedeck <- forcedeck |>
  select(`Athlete Id`, `Body Weight [kg]`, 
         `Peak Power / BM [W/kg]`, `RSI-modified [m/s]`, 
         `Countermovement Depth [cm]`, `Eccentric Braking Impulse [N s]`,
         `Concentric Duration [ms]`, 
         `Eccentric Deceleration RFD / BM [N/s/kg]`, 
         `Jump Height (Imp-Mom) [cm]`, 
         `Eccentric Deceleration RFD Asymmetry [% L,R]`, 
         `Concentric Impulse-100ms [N s]`, `Eccentric Duration [ms]`,
         `Eccentric Peak Velocity [m/s]`, `Jump Height (Flight Time) [cm]`,
         `Landing Impulse [N s]`, `Peak Landing Force / BM [N/kg]`, 
         `Landing Impulse (Asymmetry) [% L,R]`, 
         `Peak Landing Force Asymmetry [% L,R]`, 
         `Concentric Impulse (Left) [N s]`, 
         `Concentric Impulse (Right) [N s]`, 
         `Eccentric Braking Impulse (Left) [N s]`, 
         `Eccentric Braking Impulse (Right) [N s]`, 
         `Eccentric Deceleration RFD (Left) [N/s]`, 
         `Eccentric Deceleration RFD (Right) [N/s]`, 
         `Landing Impulse (Left) [N s]`, `Landing Impulse (Right) [N s]`,
         `Peak Landing Force (Left) [N]`, `Peak Landing Force (Right) [N]`)

forceframe <- forceframe |>
  select(athleteId, innerLeftMaxForce, innerRightMaxForce,
         outerLeftMaxForce, outerRightMaxForce, testPositionName)

nordbord <- nordbord |>
            select(athleteId, leftMaxForce, rightMaxForce)

perch <- perch |> 
         select(name, avg_ecc_mean_power, 
                avg_ecc_mean_velocity, 
                avg_mean_power, avg_mean_velocity, 
                avg_peak_power, avg_peak_velocity, 
                max_ecc_mean_power, max_ecc_mean_velocity,                
                max_mean_velocity, max_peak_power, 
                max_peak_velocity, pct_1rm, weight)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
plot1 <- forcedeck |> ggplot(aes(x = `RSI-modified [m/s]`)) + geom_histogram()
plot2 <- forceframe |> ggplot(aes(x = innerRightMaxForce)) + geom_histogram()
plot3 <- nordbord |> ggplot(aes(x = rightMaxForce)) + geom_histogram()
plot4 <- perch |> ggplot(aes(x = max_peak_power)) + geom_histogram()

# grid.arrange(plot1, plot2, plot3, plot4, ncol=2, nrow =2)

```
```{r message=FALSE, warning=FALSE, include=FALSE}
# Get average scores for each athlete for each exercise (removing NAs)
avg_forcedeck <- forcedeck |> 
  group_by(`Athlete Id`) |>
  summarise(across(everything(), mean, na.rm = TRUE))

avg_forceframe <- forceframe |> 
  group_by(athleteId, testPositionName) |>
  summarise(across(everything(), mean, na.rm = TRUE))
  
avg_nordboard <- nordbord |> 
  group_by(athleteId) |>
  summarise(across(everything(), mean, na.rm = TRUE))

avg_perch <- perch |> 
  select(-pct_1rm) |> 
  group_by(name) |>
  summarise(across(everything(), mean, na.rm = TRUE))
```

```{r include=FALSE}
# Join average data sets
data <- left_join(avg_nordboard, avg_forceframe, by = join_by(athleteId)) 
data <- left_join(data, avg_forcedeck, by = join_by(athleteId == `Athlete Id`))
data <- left_join(data, avg_perch, join_by(athleteId == name))
```

```{r include=FALSE}
# Consolidate body weight variable (in kg)
data <- data |>
  mutate(body_weight = case_when(
    !is.na(`Body Weight [kg]`) ~ `Body Weight [kg]`,
    !is.na(weight) ~ weight * 0.453592,
    TRUE ~ NA_real_
  )) |>
  select(-c(`Body Weight [kg]`, weight))
```

```{r include=FALSE}
# Count tests within avg_forceframe
avg_forceframe %>% ungroup() %>% count(testPositionName)

# Filter for two tests that are done the most
data <- filter(data, testPositionName == "Shoulder IR/ER - Custom" | 
                            testPositionName == "Hip AD/AB - 60")
```

```{r include=FALSE}
# Pivot testPositionName and associated four MaxForce variables
data <- data |>
  pivot_wider(names_from = testPositionName,
              values_from = c(innerLeftMaxForce, innerRightMaxForce, 
                              outerLeftMaxForce, outerRightMaxForce))
```

```{r include=FALSE}
# Calculate Asymmetries % for L/R variables
data <- data %>% 
  mutate(maxforce_asymmetry = 100 * abs((leftMaxForce - rightMaxForce)/max(leftMaxForce, rightMaxForce)), 
         innerMaxForce_Asymmetry_hips = 100 * abs((`innerLeftMaxForce_Hip AD/AB - 60` - `innerRightMaxForce_Hip AD/AB - 60`)/ pmax(`innerLeftMaxForce_Hip AD/AB - 60`, `innerRightMaxForce_Hip AD/AB - 60`)), 
         outerMaxForce_Asymmetry_hips = 100 * abs((`outerLeftMaxForce_Hip AD/AB - 60` - `outerRightMaxForce_Hip AD/AB - 60`)/ pmax(`outerLeftMaxForce_Hip AD/AB - 60`, `outerRightMaxForce_Hip AD/AB - 60`)),
         innerMaxForce_Asymmetry_shoulders = 100 * abs((`innerLeftMaxForce_Shoulder IR/ER - Custom` - `innerRightMaxForce_Shoulder IR/ER - Custom`)/ pmax(`innerLeftMaxForce_Shoulder IR/ER - Custom`, `innerRightMaxForce_Shoulder IR/ER - Custom`)),
         outerMaxForce_Asymmetry_shoulders = 100 * abs((`outerLeftMaxForce_Shoulder IR/ER - Custom` - `outerRightMaxForce_Shoulder IR/ER - Custom`)/ pmax(`outerLeftMaxForce_Shoulder IR/ER - Custom`, `outerRightMaxForce_Shoulder IR/ER - Custom`)))
```

```{r include=FALSE}
# Scale all variables except athlete ID and asymmetry %
scaled_data <- data |>
  mutate(
    leftMaxForce = scale(leftMaxForce),
    rightMaxForce = scale(rightMaxForce),
    `Peak Power / BM [W/kg]` = scale(`Peak Power / BM [W/kg]`),
    `RSI-modified [m/s]` = scale(`RSI-modified [m/s]`),
    `Countermovement Depth [cm]` = scale(`Countermovement Depth [cm]`),
    `Eccentric Braking Impulse [N s]` = scale(`Eccentric Braking Impulse [N s]`),
    `Concentric Duration [ms]` = scale(`Concentric Duration [ms]`),
    `Eccentric Deceleration RFD / BM [N/s/kg]` = scale(`Eccentric Deceleration RFD / BM [N/s/kg]`),
    `Jump Height (Imp-Mom) [cm]` = scale(`Jump Height (Imp-Mom) [cm]`),
    `Concentric Impulse-100ms [N s]` = scale(`Concentric Impulse-100ms [N s]`),
    `Eccentric Duration [ms]` = scale(`Eccentric Duration [ms]`),
    `Eccentric Peak Velocity [m/s]` = scale(`Eccentric Peak Velocity [m/s]`),
    `Jump Height (Flight Time) [cm]` = scale(`Jump Height (Flight Time) [cm]`),
    `Landing Impulse [N s]` = scale(`Landing Impulse [N s]`),
    `Peak Landing Force / BM [N/kg]` = scale(`Peak Landing Force / BM [N/kg]`),
    `Concentric Impulse (Left) [N s]` = scale(`Concentric Impulse (Left) [N s]`),
    `Concentric Impulse (Right) [N s]` = scale(`Concentric Impulse (Right) [N s]`),
    `Eccentric Braking Impulse (Left) [N s]` = scale(`Eccentric Braking Impulse (Left) [N s]`),
    `Eccentric Braking Impulse (Right) [N s]` = scale(`Eccentric Braking Impulse (Right) [N s]`),
    `Eccentric Deceleration RFD (Left) [N/s]` = scale(`Eccentric Deceleration RFD (Left) [N/s]`),
    `Eccentric Deceleration RFD (Right) [N/s]` = scale(`Eccentric Deceleration RFD (Right) [N/s]`),
    `Landing Impulse (Left) [N s]` = scale(`Landing Impulse (Left) [N s]`),
    `Landing Impulse (Right) [N s]` = scale(`Landing Impulse (Right) [N s]`),
    `Peak Landing Force (Left) [N]` = scale(`Peak Landing Force (Left) [N]`),
    `Peak Landing Force (Right) [N]` = scale(`Peak Landing Force (Right) [N]`),
    avg_ecc_mean_power = scale(avg_ecc_mean_power),
    avg_ecc_mean_velocity = scale(avg_ecc_mean_velocity),
    avg_mean_power = scale(avg_mean_power),
    avg_mean_velocity = scale(avg_mean_velocity),
    avg_peak_power = scale(avg_peak_power),
    avg_peak_velocity = scale(avg_peak_velocity),
    max_ecc_mean_power = scale(max_ecc_mean_power),
    max_ecc_mean_velocity = scale(max_ecc_mean_velocity),
    max_mean_velocity = scale(max_mean_velocity),
    max_peak_power = scale(max_peak_power),
    max_peak_velocity = scale(max_peak_velocity),
    body_weight = scale(body_weight),
    `innerLeftMaxForce_Hip AD/AB - 60` = scale(`innerLeftMaxForce_Hip AD/AB - 60`),
    `innerLeftMaxForce_Shoulder IR/ER - Custom` = scale(`innerLeftMaxForce_Shoulder IR/ER - Custom`),
    `innerRightMaxForce_Hip AD/AB - 60` = scale(`innerRightMaxForce_Hip AD/AB - 60`),
    `innerRightMaxForce_Shoulder IR/ER - Custom` = scale(`innerRightMaxForce_Shoulder IR/ER - Custom`),
    `outerLeftMaxForce_Hip AD/AB - 60` = scale(`outerLeftMaxForce_Hip AD/AB - 60`),
    `outerLeftMaxForce_Shoulder IR/ER - Custom` = scale(`outerLeftMaxForce_Shoulder IR/ER - Custom`),
    `outerRightMaxForce_Hip AD/AB - 60` = scale(`outerRightMaxForce_Hip AD/AB - 60`),
    `outerRightMaxForce_Shoulder IR/ER - Custom` = scale(`outerRightMaxForce_Shoulder IR/ER - Custom`)
  )
```

```{r warning=FALSE, echo=FALSE, fig.width=12, fig.height=10}
data1 <- data |> select(`RSI-modified [m/s]`, rightMaxForce, max_peak_power, `innerRightMaxForce_Hip AD/AB - 60`, `outerRightMaxForce_Hip AD/AB - 60`) 

ggpairs(data1)
```


## Methodology

### Data cleaning

For each of the four instruments, we selected a single variable that is most representative of what each of the instruments intends to assess. For `forcedeck`, we selected `RSI-modified` which stands for Reactive Strength Index and is a measure of explosiveness, neuromuscular readiness and fatigue. In the case of `forceframe`, athletes performed multiple tests with this instrument. Most of the tests were performed by less than 30 athletes. Therefore, we chose to analyze performance in the two tests that were performed the most. These were: `Hip Adduction/Abduction` with a 60 degree supine position (70 athletes) and `Shoulder Internal/External Rotations` (83 athletes) with a custom setup. For the hips tests, we chose the `Inner Right Max Force` because this represents the adduction force, which is related to groin and adductor muscle strength. For the shoulders test, we chose the `Outer Right Max Force` because this is the extension force, which is crucial in football, especially linemen. Then, for the `nordbord`, we selected the `Right Max Force` as a measurement of hamstring strength. Finally, for `perch`, we selected `Max Peak Power` as a measure of explosiveness.

It is important to note the Duke Athletics Sports Science team did not express a preference for a side when metrics where available for both left and right limbs. Therefore, we chose to consistently use the right side metrics in all cases where this applied, based on the assumption that it is the dominant side for the majority of athletes. However, this choice is not expected to have any significant impact on our analysis.

To explore asymmetries, we also chose a metric from each of the four instruments. For `forcedeck`, the data already contained the following left/right asymmetry metrics in percentage of asymmetry: `Landing Impulse Asymmetry`, `Peak Landing Force Asymmetry`, and `Eccentric Deceleration RFD Asymmetry`. We calculated left/right asymmetry percentages for `Inner Max Force` and `Outer Max Force` for both the hips and shoulders tests in `forceframe` and for `Max Force` in `nordbord`. To do so, we used the following formula, which was used the calculate the asymmetry percentages already included in the the `forcedeck` dataset: $\text{Left/Right Asymmetry Percentage} = \left| \frac{\text{Left}+\text{Right}}{\max(\text{Left}, \text{Right})} \right| \times 100$. Perch did not provide any metrics from which asymmetries could be calculated.

To analyze relationships across the four datasets, we decided to average measurements for each athlete for each variable of interest. To ensure that information was not lost, we checked that variances between athletes were generally low for the variables of interest (see Appendix). Variances all appeared to be centered near zero and skewed right, with the exception of `Outer Right Max Force`.

We then combined the datasets using the athlete ID as an identifier. Since body weight was found in the `forcedeck` dataset in kilograms and was also collected in the `perch` dataset in pounds, we combined the two variables by using the kilograms value from `forcedeck` if present and if not, converting the value from `perch` to kilograms. Finally, we scaled all variables so that individual variables would not disproportionately influence the outcome of the model by standardizing each variable to a mean of 0 and variance of 1.

### Modeling for performance relationships across instruments

To explore the associations between performance across the different instruments, we fitted a set of 5 linear regression models using the scaled data previously described. We used the following variables of interest in each of the models: `Reactive Strength Index` from `forcedeck`, `Inner Right Max Force` from `forceframe` hip tests, `Outer Right Max Force` from `forceframe` shoulder tests, `Right Max Force` from `nordbord`, and `Max Peak Power` from `perch.` In the 5 models, each of these variables served as the response variable while the other four where included as predictors. Moreover, body weight was included in each of the models to account for variability of athlete weights within a football team. Linear model diagnostics were assessed and can be found in the Appendix.

```{r include=FALSE}
# Forcedeck Model *
fd_mod <- lm(`RSI-modified [m/s]` ~ 
               rightMaxForce + max_peak_power +
               `innerRightMaxForce_Hip AD/AB - 60` + 
               `outerRightMaxForce_Shoulder IR/ER - Custom` + 
               body_weight, data = scaled_data)
```

```{r include=FALSE}
# Forceframe Hip Model
ff_mod_hip <- lm(`innerRightMaxForce_Hip AD/AB - 60` ~
                `RSI-modified [m/s]` + rightMaxForce + 
                  max_peak_power + 
                  `outerRightMaxForce_Shoulder IR/ER - Custom` + 
                  body_weight, data = scaled_data)
```

```{r include=FALSE}
# Force frame shoulder model *
ff_mod_shoulder <- lm(`outerRightMaxForce_Shoulder IR/ER - Custom` ~
                `RSI-modified [m/s]` + rightMaxForce + 
                  max_peak_power + `innerRightMaxForce_Hip AD/AB - 60` + 
                  body_weight, 
                data = scaled_data)
```

```{r include = FALSE}
# Nordbord model * 
nb_mod <- lm(rightMaxForce ~
                `RSI-modified [m/s]` + 
               `outerRightMaxForce_Shoulder IR/ER - Custom` + 
                  max_peak_power +
                `innerRightMaxForce_Hip AD/AB - 60` + body_weight, data = scaled_data)
```

```{r include = FALSE}
# Perch model *
p_mod <- lm(max_peak_power ~
               `RSI-modified [m/s]` + `outerRightMaxForce_Shoulder IR/ER - Custom` + 
                 rightMaxForce+
               `innerRightMaxForce_Hip AD/AB - 60` + body_weight, data = scaled_data)
```

### Modeling for the effect of asymmetries in performance of other instruments

To investigate the associations of left/right asymmetries on performance across the different instruments, we conducted another set of 5 linear regression models. The goal of these models was to evaluate the hypothesis that higher asymmetry percentages have a negative association on performance and to identify which asymmetric body parts have these negative effects. We fitted a model for each of the 5 variables we defined to be representatives of performance in the 4 instruments. These target variables are: `Reactive Strength Index` from `forcedeck`, `Inner Right Max Force` from `forceframe` hip tests, `Outer Right Max Force` from `forceframe` shoulder tests, `Right Max Force` from `nordbord`, and `Max Peak Power` from `perch.` The predictors for these five models were all left/right asymmetry percentages. More specifically, these were: `Landing Impulse Asymmetry`, `Peak Landing Force Asymmetry`, and `Eccentric Deceleration RFD Asymmetry` from `forcedeck`; `Inner Max Force Asymmetry` and `Outer Max Force Asymmetry` for both hips and shoulders tests from `forceframe`; and `Max Force Asymmetry` from `nordbord.` In each of the models we only included predictors that came from a different instrument that the one that measured the target variable. Body weight was included in all models to account for the variability across athletes in within the team. Linear model diagnostics were checked and can be found in the Appendix.

```{r include = FALSE}
# Forcedeck Asymmetry Model
fd_asy_mod <- lm( `RSI-modified [m/s]` ~ maxforce_asymmetry + 
                innerMaxForce_Asymmetry_hips + 
                innerMaxForce_Asymmetry_shoulders + 
                outerMaxForce_Asymmetry_hips + 
                outerMaxForce_Asymmetry_shoulders + body_weight, data = scaled_data)
```

```{r include = FALSE}
# Forceframe Hip Asymmetry Model * 
ff_asy_hip_mod <- lm(`innerRightMaxForce_Hip AD/AB - 60` ~ 
             `Landing Impulse (Asymmetry) [% L,R]`+
                `Peak Landing Force Asymmetry [% L,R]`+
                `Eccentric Deceleration RFD Asymmetry [% L,R]`+ 
                maxforce_asymmetry + 
                innerMaxForce_Asymmetry_shoulders + 
                outerMaxForce_Asymmetry_shoulders + body_weight, data = scaled_data)
```

```{r include = FALSE}
# Forceframe Shoulder Asymmetry Model
ff_asym_shoulder_mod <- lm(`outerRightMaxForce_Shoulder IR/ER - Custom` ~ 
             `Landing Impulse (Asymmetry) [% L,R]`+
                `Peak Landing Force Asymmetry [% L,R]`+
                `Eccentric Deceleration RFD Asymmetry [% L,R]`+ 
                maxforce_asymmetry + 
                innerMaxForce_Asymmetry_hips + 
                outerMaxForce_Asymmetry_hips + body_weight, data = scaled_data)
```

```{r include = FALSE}
# Nordbord Asymmetry Model
nb_asy_mod <- lm(rightMaxForce ~ `Landing Impulse (Asymmetry) [% L,R]`+
                `Peak Landing Force Asymmetry [% L,R]`+
                `Eccentric Deceleration RFD Asymmetry [% L,R]`+ 
                innerMaxForce_Asymmetry_hips + 
                innerMaxForce_Asymmetry_shoulders + 
                outerMaxForce_Asymmetry_hips + 
                outerMaxForce_Asymmetry_shoulders + body_weight, data = scaled_data)
```

```{r include = FALSE}
# Perch Asymmetry Model:
p_asym_model <- lm( max_peak_power ~ `Landing Impulse (Asymmetry) [% L,R]`+
                `Peak Landing Force Asymmetry [% L,R]`+
                `Eccentric Deceleration RFD Asymmetry [% L,R]`+ 
                maxforce_asymmetry + 
                innerMaxForce_Asymmetry_hips + 
                innerMaxForce_Asymmetry_shoulders + 
                outerMaxForce_Asymmetry_hips + 
                outerMaxForce_Asymmetry_shoulders + body_weight, data = scaled_data)
```

## Results

### Performance Models

Max Peak Power from perch appeared to have a significant positive association with performance on each of the other instruments. More specifically, Reactive Strength Index was significantly associated (p=0.009) with Max Peak Power. 
We expect a 0.338 unit increase in Reactive Strength Index for every one unit increase in Max Peak Power, holding constant all other predictors in the model. This indicates that athletes who generated a higher peak power during a squad also showed more explosiveness, neuromuscular readiness and less fatigue during their vertical jumps. The variables included from nordbord and forceframe were not statistically significant. 

In the case of force deck, only Shoulder Outer Right Max Force was significantly associated (p=0.018) with Max Peak Power but Hip Inner Right Max Force was not. For every one unit increase in Max Peak Power, we expect a 0.319 unit increase in Shoulder Outer Right Max Force, holding constant all other predictors in the model. This suggests that players who generated higher peak power during a squad also generated more outer force in their shoulders. Variables from the other instruments included in the model were not significant. 

Finally, Right Max Force from nordbord was also significantly associated (p=0.045) with  Max Peak Power. For every one unit increase in Max Peak Power, we expect a 0.301 unit increase in Max Right Force, e expect a 0.319 unit increase in Shoulder Outer Right Max Force. This means that athletes who generated higher peak power during a squad also generated higher force with their hamstrings. Predictors from other instruments were not statistically significant. 

All of these results were evaluated at the 0.05 alpha level.

```{r echo=FALSE}
fd_mod_summary <- summary(fd_mod)

rownames(fd_mod_summary$coefficients) <- c("Intercept","Right Max Force (nordbord)", "Max Peak Power (perch)", "Inner Right Max Force Hip AD/AB (forceframe)",  "Outer Right Max Force Shoulder IR/ER (forceframe)", "Body Weight")

coeffs_rounded <- round(fd_mod_summary$coefficients, 3)
coeffs_rounded[, 4] <- ifelse(coeffs_rounded[, 4] < 0.001, "<0.001", coeffs_rounded[, 4])

kable(coeffs_rounded, 
      caption = "Linear Regression Model for RSI-modified (forcedeck)", 
      col.names = c("Coefficient", "Estimate", "Std. Error", "t-value", "p-value"))
```


```{r echo=FALSE}
ff_mod_shoulder_summary <- summary(ff_mod_shoulder)

rownames(ff_mod_shoulder_summary$coefficients) <- c("Intercept", "RSI modified (forcedeck)", "Right Max Force (nordbord)", "Max Peak Power (perch)",  "Inner Right Max Force Hip AD/AB (forceframe)", "Body Weight")

kable(round(ff_mod_shoulder_summary$coefficients, 3),
      caption = "Linear Regression Model for Outer Right Max Force Shoulder IR/ER - Custom (forceframe)", 
      col.names = c("Coefficient", "Estimate", "Std. Error", "t-value", "p-value"))
```

```{r echo=FALSE}
nb_mod_summary <- summary(nb_mod)

rownames(nb_mod_summary$coefficients) <- c("Intercept", "RSI modified (forcedeck)", "Outer Right Max Force Shoulder IR/ER (forceframe)", "Max Peak Power (perch)",  "Inner Right Max Force Hip AD/AB (forceframe)", "Body Weight")

kable(round(nb_mod_summary$coefficients, 3),
      caption = "Linear Regression Model for Right Max Force (nordbord)", 
      col.names = c("Coefficient", "Estimate", "Std. Error", "t-value", "p-value"))
```

```{r echo=FALSE}
p_mod_summary <- summary(p_mod)

rownames(p_mod_summary$coefficients) <- c("Intercept", "RSI modified (forcedeck)", "Outer Right Max Force Shoulder IR/ER (forceframe)", "Right Max Force (nordbord)",  "Inner Right Max Force Hip AD/AB (forceframe)", "Body Weight")

kable(round(p_mod_summary$coefficients, 3),
      caption = "Linear Regression Model for Max Peak Power (perch)", 
      col.names = c("Coefficient", "Estimate", "Std. Error", "t-value", "p-value"))
```

### Asymmetry Models

From the five asymmetry percentage models, the only significant association (p=0.045) was between Hip Inner Right Max Force and Max Force Asymmetry Percentage from nordbord. For every one percentage point increase in Max Force Asymmetry Percentage, we expect Hip Inner Right Max Force to decrease by -0.040 units, holding constant all other predictors in the model. This suggest a negative association between weaknesses or imbalances in an athletes hamstring and their capacity to generate force in their hips. This result was evaluated at the 0.05 alpha level.

```{r echo=FALSE}
ff_asy_hip_mod_summary <- summary(ff_asy_hip_mod)

rownames(ff_asy_hip_mod_summary$coefficients) <- c("Intercept", "Landing Impulse Asymmetry % (forcedeck)", "Peak Landing Force Asymmetry % (forcedeck)", "Eccentric Deceleration RFD Asymmetry % (forcedeck)" , "Max Force Asymmetry % (nordbord)", "Shoulders Inner Max Force Asymmetry % (forceframe)", "Shoulders Outer Max Force Asymmetry % (forceframe)", "Body Weight")

coeffs_rounded <- round(ff_asy_hip_mod_summary$coefficients, 3)
coeffs_rounded[, 4] <- ifelse(coeffs_rounded[, 4] < 0.001, "<0.001", coeffs_rounded[, 4])

kable(coeffs_rounded,
      caption = "Asymmetry Linear Regression Model for Inner Right Max Force Hip AD/AB - 60 (forceframe)", 
      col.names = c("Coefficient", "Estimate", "Std. Error", "t-value", "p-value"))
```

## Discussion

The first important conclusion from these models is that the `Max Peak Power` metric from the perch exercise is significantly associated with the metrics of interest from the other variables. This is noteworthy as many of the other metrics of interest do not have this association among themselves, but `Max Peak Power` does. This could mean that the perch exercise should be a point of focus for coaches and athletic trainers as it can act as an indicator for strength, power, neuromuscular readiness, and fatigue in the other exercises.

The second main conclusion is from the asymmetry models. These models were fit in an effort to evaluate if strength asymmetries in athletes impact just the relevant exercise or others. Do these imbalances correlate to reduced scores in other metrics? Our models show that those imbalances do not show correlation with strength scores from other parts of the body other than with hamstrings as tested through the nordbord. The fact that hamstring and hip force are negatively correlated may indicate a trade off between these systems of the body. This may also be interesting to Duke football coaches and athletic trainers in terms of injury recovery and prevention and strength training focus.

There are of course limits to this analysis, as we are not members of the community that have full access to the data. Without data like player position, injuries, training focus for certain periods of time, we do not have a complete picture. Data errors from mis-reads to missed tests have also complicated the analysis. Moreover, out data and results are only indicative of performance between these instruments but not on-field performance, which is the ultimate goal. Also, our data does not account for the players load differences (startes play and are more tired than players than see less minutes) which might be another factor to consider. Our approach is simplistic but effective, some sort of hierarchical approach could be a future direction for this analysis, relying on athletes within position groups within teams, allowing us to look across schools to find patterns in the general population. 


## Appendix

### Linear model diagnostics

Linear Regression Model for RSI-modified (forcedeck):  

```{r echo=FALSE}
#| fig-width: 6
#| fig-height: 3

plot(fd_mod)
```


\newpage
Linear Regression Model for Outer Right Max Force Shoulder IR/ER - Custom (forceframe):

```{r echo=FALSE}
#| fig-width: 6
#| fig-height: 3


plot(ff_mod_shoulder)
```

\newpage
Linear Regression Model for Right Max Force (nordbord):

```{r echo=FALSE}
#| fig-width: 6
#| fig-height: 3


plot(nb_mod)
```

\newpage
Linear Regression Model for Max Peak Power (perch):

```{r echo=FALSE}
#| fig-width: 6
#| fig-height: 3

plot(p_mod)
```

\newpage
Asymmetry Linear Regression Model for Inner Right Max Force Hip AD/AB - 60 (forceframe):

```{r echo=FALSE}
#| fig-width: 6
#| fig-height: 3

plot(ff_asy_hip_mod)
```

\newpage

### Variance between athletes

To ensure that the aggregation of data by athlete was appropriate, we checked the variance between athletes to ensure that there was minimal variance and data was not lost.

```{r echo=FALSE}
#| fig-width: 6
#| fig-height: 3

forcedeck_vars <- c(
  "Athlete Id",
  "RSI-modified [m/s]",
  "Peak Landing Force (Left) [N]",
  "Peak Landing Force (Right) [N]",
  "Landing Impulse (Left) [N s]",
  "Landing Impulse (Right) [N s]"
)

forcedeck_long <- forcedeck |>
  select(all_of(forcedeck_vars)) |>
  pivot_longer(-`Athlete Id`, names_to = "variable", values_to = "value") |>
  group_by(`Athlete Id`, variable) |>
  summarise(var = var(value, na.rm = TRUE), .groups = "drop")

ggplot(forcedeck_long, aes(x = var)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

```{r warning=FALSE, echo=FALSE}
#| fig-width: 6
#| fig-height: 3

forceframe_vars <- c(
  "athleteId",
  "innerRightMaxForce",
  "innerLeftMaxForce",
  "outerRightMaxForce",
  "outerLeftMaxForce"
)

forceframe_long <- forceframe |>
  select(all_of(forceframe_vars)) |>
  pivot_longer(-athleteId, names_to = "variable", values_to = "value") |>
  group_by(athleteId, variable) |>
  summarise(var = var(value, na.rm = TRUE), .groups = "drop")

ggplot(forceframe_long, aes(x = var)) +
  geom_histogram(bins = 30, fill = "tomato", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

```{r warning=FALSE, echo=FALSE}
#| fig-width: 6
#| fig-height: 3

nordbord_vars <- c(
  "athleteId",
  "rightMaxForce",
  "leftMaxForce"
)

nordbord_long <- nordbord |>
  select(all_of(nordbord_vars)) |>
  pivot_longer(-athleteId, names_to = "variable", values_to = "value") |>
  group_by(athleteId, variable) |>
  summarise(var = var(value, na.rm = TRUE), .groups = "drop")

ggplot(nordbord_long, aes(x = var)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

```{r echo=FALSE}
#| fig-width: 6
#| fig-height: 3

perch_vars <- c(
  "name",
  "max_peak_power"
)

perch_long <- perch |>
  rename(`Max Peak Power` = max_peak_power) |>
  select(name, `Max Peak Power`) |>
  pivot_longer(-name, names_to = "variable", values_to = "value") |>
  group_by(name, variable) |>
  summarise(var = var(value, na.rm = TRUE), .groups = "drop")

ggplot(perch_long, aes(x = var)) +
  geom_histogram(bins = 30, fill = "purple", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

